Accelerating Neural Network Training using Arbitrary Precision Approximating Matrix Multiplication Algorithms

### 2021-03-05 latex/MPP21/paper.tex
set up NN overall reltime plots
added description of householder, inserted speedup numbers in abstract and intro

\begin{abstract}
Matrix multiplication is one of the bottleneck computations for training the weights within deep neural networks.
To speed up the training phase, we propose to use faster algorithms for matrix multiplication known as Arbitrary Precision Approximating (APA) algorithms.
APA algorithms perform asymptotically fewer arithmetic operations than the classical algorithm, but they compute an approximate result with an error that can be made arbitrarily small in exact arithmetic.
Practical APA algorithms provide significant reduction in computation time and still provide enough accuracy for many applications like neural network training.
We demonstrate that APA algorithms can be efficiently implemented and parallelized for shared-memory platforms to obtain up to \GB{Z\%} speedups over the fastest implementation of the classical algorithm.
Furthermore, using these algorithms to train a Multi-Layer Perceptron (MLP) network yields no signification reduction in the training or testing error.
Our performance results on a large MLP network show overall performance improvements of up to \GB{W\%}.
\end{abstract}

@@

Our contribution is the use of APA matrix multiplication algorithms to address this problem and accelerate training.
In particular we
\begin{enumerate}
	\item curate a collection of well-established and recently discovered practical APA algorithms;
	\item extend the framework of \cite{BB15} to generate efficient multithreaded code for all of them;
	\item demonstrate the robustness of learning accuracy to approximate matrix multiplications; and
	\item present multithreaded performance improvements of a synthetic MLP of up to \GB{$Y\times$} over the use of classical matrix multiplication.
\end{enumerate}

\GB{\cite{NPOV15} uses a network including layers with dimensions 25K x 4K and 4K x 4K}

@@

A key characteristic of matrix multiplication, first demonstrated by Bini et al.~\cite{BCRL79}, is that it can be approximated to arbitrary accuracy more efficiently than being computed exactly.
An APA algorithm is one that takes as input $A$, $B$, and a parameter $0<\lambda<1$ and computes 
\begin{equation}
\label{eq:APAapprox}
\hat C = A\cdot B + \lambda E + O(\lambda^2).
\end{equation}
Thus, in exact arithmetic, letting $\lambda \rightarrow 0$ achieves arbitrarily small approximation error.
In floating point arithmetic, choosing too small a value for $\lambda$ leads to accumulation of roundoff error that exceeds the approximation error.
We discuss optimizing $\lambda$ in \cref{sec:APAerr}.

@@

For a concrete example, we reproduce the rule for the algorithm developed by Bini et al.~\cite{BCRL79} for the $\dims{3,2,2}$ case (multiplying a $3\times 2$ matrix $A$ by a $2\times 2$ matrix $B$), where we use the following notation for input and output matrices:
$$\begin{bmatrix} C_{11} & C_{12} \\ C_{21} & C_{22} \\ C_{31} & C_{32} \end{bmatrix} = 
\begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \\ A_{31} & A_{32}  \end{bmatrix} \cdot
\begin{bmatrix} B_{11} & B_{12} \\ B_{21} & B_{22} \end{bmatrix}.$$
Bini's rule is given by
\begin{align*}
M_1 =& (A_{11} + A_{22}) \cdot (\lambda B_{11} + B_{22}) \\
M_2 =& A_{22}\cdot  (-B_{21} - B_{22}) \\
M_3 =& A_{11}\cdot B_{22} \\
M_4 =& (\lambda A_{12} + A_{22})\cdot (-\lambda B_{11} + B_{21}) \\
M_5 =& (A_{11} + \lambda A_{12}) \cdot (\lambda B_{12} + B_{22}) \\
M_6 =& (A_{21} + A_{32}) \cdot (B_{11} + \lambda B_{22}) \\
M_7 =& A_{21} \cdot (-B_{11} - B_{12}) \\
M_8 =& A_{32} \cdot B_{11}\\
M_9 =& (A_{21} + \lambda A_{31}) \cdot (B_{12} - \lambda B_{22})\\
M_{10} =& (\lambda A_{31} + A_{32}) \cdot (B_{12} - \lambda B_{22})\\
\end{align*}
\begin{align*}
C_{11} =& \lambda^{-1} (M_1 + M_2 - M_3 + M_4) \\
C_{12} =& \lambda^{-1} (-M_3 + M_5) \\
C_{21} =& M_4 + M_6 - M_{10}\\
C_{22} =& M_1 - M_5 + M_9\\
C_{31} =& \lambda^{-1} (-M_8 + M_{10}) \\
C_{32} =& \lambda^{-1} (M_6 + M_7 - M_8 + M_9).\\
\end{align*}

@@

@@

\GB{add paragraph describing householder, listing versions of MKL, peak GFLOPS}

@@

\begin{figure}
\subfloat[One thread]{
\centering
\renewcommand{\datafile}{data/NNtime_1t.dat}
\input{plots/reltime}
\label{fig:NNreltime_seq}}
\\
\subfloat[Twelve threads]{
\centering
\renewcommand{\datafile}{data/NNtime_12t.dat}
\input{plots/reltime}
\label{fig:NNreltime_par}}
\caption{Network training time relative to using classical matrix multiplication}
\label{fig:NNreltime}
\end{figure}

@@

\begin{enumerate}
	\item promising results for synthetic network, goal is to target costly networks bottlenecked by matmul
	\item can tune for matmul dimensions (not just square)
	\item GPU implementation (cite Jianyu Huang's work)
	\item mixed precision fast matmul (APA or otherwise)
\end{enumerate}

>>>>

\begin{abstract}
Matrix multiplication is one of the bottleneck computations for training the weights within deep neural networks.
To speed up the training phase, we propose to use faster algorithms for matrix multiplication known as Arbitrary Precision Approximating (APA) algorithms.
APA algorithms perform asymptotically fewer arithmetic operations than the classical algorithm, but they compute an approximate result with an error that can be made arbitrarily small in exact arithmetic.
Practical APA algorithms provide significant reduction in computation time and still provide enough accuracy for many applications like neural network training.
We demonstrate that APA algorithms can be efficiently implemented and parallelized for shared-memory platforms to obtain up to 28\% and 21\% speedups over the fastest implementation of the classical algorithm using one core and 12 cores, respectively.
Furthermore, using these algorithms to train a Multi-Layer Perceptron (MLP) network yields no signification reduction in the training or testing error.
Our performance results on a large MLP network show overall performance improvements of up to \GB{W\%}.
\end{abstract}

@@

Our contribution is the use of APA matrix multiplication algorithms to address this problem and accelerate training.
In particular we
\begin{enumerate}
	\item curate a collection of well-established and recently discovered practical APA algorithms;
	\item extend the framework of \cite{BB15} to generate efficient multithreaded code for all of them, achieving up to 21\% performance improvement over the best parallel classical implementation,
	\item demonstrate the robustness of learning accuracy to approximate matrix multiplications; and
	\item present multithreaded performance improvements of a synthetic MLP of up to \GB{$Y\times$} over the use of classical matrix multiplication.
\end{enumerate}

@@

A key characteristic of matrix multiplication, first demonstrated by Bini et al.~\cite{BCRL79}, is that it can be approximated to arbitrary accuracy more efficiently than being computed exactly.
An APA algorithm is one that takes as input $A$, $B$, and a scalar parameter $0<\lambda<1$ and computes 
\begin{equation}
\label{eq:APAapprox}
\hat C = A\cdot B + \lambda E + O(\lambda^2).
\end{equation}
Thus, in exact arithmetic, letting $\lambda \rightarrow 0$ achieves arbitrarily small approximation error.
In floating point arithmetic, choosing too small a value for $\lambda$ leads to accumulation of roundoff error that exceeds the approximation error.
We discuss optimizing $\lambda$ in \cref{sec:APAerr}.

@@

For a concrete example, we reproduce the rule for the algorithm developed by Bini et al.~\cite{BCRL79} for the $\dims{3,2,2}$ case (multiplying a $3\times 2$ matrix $A$ by a $2\times 2$ matrix $B$), where we use the following notation for input and output matrices:
$$
\begin{bmatrix} A_{11} & A_{12} \\ A_{21} & A_{22} \\ A_{31} & A_{32}  \end{bmatrix} \cdot
\begin{bmatrix} B_{11} & B_{12} \\ B_{21} & B_{22} \end{bmatrix}=
\begin{bmatrix} C_{11} & C_{12} \\ C_{21} & C_{22} \\ C_{31} & C_{32} \end{bmatrix} .$$
Bini's rule is given by
\begin{align*}
M_1 =& (A_{11} + A_{22}) \cdot (\lambda B_{11} + B_{22}) \\
M_2 =& A_{22}\cdot  (-B_{21} - B_{22}) \\
M_3 =& A_{11}\cdot B_{22} \\
M_4 =& (\lambda A_{12} + A_{22})\cdot (-\lambda B_{11} + B_{21}) \\
M_5 =& (A_{11} + \lambda A_{12}) \cdot (\lambda B_{12} + B_{22}) \\
M_6 =& (A_{21} + A_{32}) \cdot (B_{11} + \lambda B_{22}) \\
M_7 =& A_{21} \cdot (-B_{11} - B_{12}) \\
M_8 =& A_{32} \cdot B_{11}\\
M_9 =& (A_{21} + \lambda A_{31}) \cdot (B_{12} - \lambda B_{22})\\
M_{10} =& (\lambda A_{31} + A_{32}) \cdot (B_{12} - \lambda B_{22})\\
\end{align*}
\begin{align*}
C_{11} =& \lambda^{-1} (M_1 + M_2 - M_3 + M_4) \\
C_{12} =& \lambda^{-1} (-M_3 + M_5) \\
C_{21} =& M_4 + M_6 - M_{10}\\
C_{22} =& M_1 - M_5 + M_9\\
C_{31} =& \lambda^{-1} (-M_8 + M_{10}) \\
C_{32} =& \lambda^{-1} (M_6 + M_7 - M_8 + M_9).\\
\end{align*}

@@

All experiments are performed on a dual-socket Intel Xeon E5-2620 (Sandy Bridge) server with 2 sockets each with 6 cores. 
Each socket has a 15 MB L3 cache, and each core has a 256 KB L2 cache and 32 KB L1 data cache. 
Each core has a clock rate of 2.00 GHz (with turbo boost disabled) and peak single precision flop rate of 32 GFLOPS. 
Our code is compiled with GCC version 7.5.0, 
and we use Intel's Math Kernel Library (MKL) version 2019.4.243.

@@

@@

\begin{figure}
\subfloat[One thread]{
\centering
\renewcommand{\datafile}{data/NN_seq.dat}
\input{plots/NN_reltime}
\label{fig:NNreltime_seq}}
\\
\subfloat[Twelve threads]{
\centering
\renewcommand{\datafile}{data/NNtime_12t.dat}
%\input{plots/NN_reltime}
\label{fig:NNreltime_par}}
\caption{Network training time relative to using classical matrix multiplication}
\label{fig:NNreltime}
\end{figure}

@@

\begin{enumerate}
	\item promising results for synthetic network, goal is to target costly networks bottlenecked by matmul (\GB{\cite{NPOV15} uses a network including layers with dimensions 25K x 4K and 4K x 4K})
	\item can tune for matmul dimensions (not just square)
	\item GPU implementation (cite Jianyu Huang's work)
	\item mixed precision fast matmul (APA or otherwise)
\end{enumerate}
